---
title: "Getting started with dryaddata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dryaddata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Chunks below hit live Dryad. They evaluate only when the build opts in
# via `Sys.setenv(DRYADDATA_BUILD_LIVE_VIGNETTES = "true")`; otherwise the
# vignette knits as static text. See `?dryaddata::dryad_vignette_live`.
live <- dryaddata::dryad_vignette_live()
has_credentials <- live && dryaddata::dryad_has_credentials()
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = live)
```

`dryaddata` is a thin, modern R client for the Dryad REST API (v2). It
helps you search and retrieve dataset metadata, walk through versions and
files, and download content. OAuth2 authentication and a shared on-disk
cache are built in.

```{r}
library(dryaddata)
```

## Production by default

`dryaddata` is pointed at Dryad's production deployment out of the box:

```{r}
dryad_base_url()
#> [1] "https://datadryad.org/api/v2"
```

To register for OAuth2 credentials (needed for the `dryad_download_*()`
family), follow the setup steps in
`vignette("downloading", package = "dryaddata")`.

If you want to experiment without touching production data, `dryaddata`
also supports Dryad's public sandbox. `dryad_use_sandbox()` repoints both
the API and the OAuth token URLs at the sandbox host; `dryad_use_sandbox(FALSE)`
reverses the change. The helper returns the previous URLs invisibly so you
can restore them. Sandbox accounts are independent of production; register
separately at <https://sandbox.datadryad.org/account>.

## Anatomy of a list response

Endpoints that return lists (search, datasets, versions, files) all share
a common envelope:

```{r}
hits <- dryad_search(q = "soil", per_page = 5)

names(hits)
#> [1] "metadata" "records"  "data"

hits$metadata$count   # records on this page
hits$metadata$total   # total matches across all pages
head(hits$data[, c("identifier", "title")])
```

* `metadata` holds counts and the raw HAL `_links` block (`self`, `next`,
  `prev`, ...).
* `records` is the untouched list of objects as returned by the API.
  Useful when you need a nested field like `authors` or `_links`.
* `data` is a tidy data frame that stacks scalar fields and keeps
  nested ones as list-columns.

## Walking from a DOI to bytes

```{r}
ds <- dryad_dataset("doi:10.5061/dryad.j1fd7")
ds$title
ds$abstract

vs <- dryad_dataset_versions(ds$identifier)
latest <- vs$data$id[which.max(vs$data$versionNumber)]

fl <- dryad_version_files(latest)
fl$data[, c("path", "size", "mimeType")]
```

Downloading a file requires authenticated credentials (see
`?dryad_auth`). The chunk below only runs when `DRYAD_CLIENT_ID` and
`DRYAD_CLIENT_SECRET` are set in the build environment.

Pass `path =` to set the destination file; otherwise the file is
cached at `dryad_cache_dir()`. The example uses `tempdir()` so the
downloaded file is removed when the R session ends.

```{r, eval = has_credentials}
csv_path <- dryad_download_file(
  fl$data$id[1],
  path = file.path(tempdir(), "first-file.csv")
)
csv_path
```

## Where to go next

* Pagination, filters, and the search query language:
  `vignette("searching")`.
* OAuth setup, the download cache, and large transfers:
  `vignette("downloading")`.
