---
title: "Searching Dryad"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Searching Dryad}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# See `?dryaddata::dryad_vignette_live`: live chunks run only when the build
# opts in via `Sys.setenv(DRYADDATA_BUILD_LIVE_VIGNETTES = "true")`.
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = dryaddata::dryad_vignette_live()
)
```

`dryad_search()` wraps `GET /search`. The package targets Dryad's
production deployment by default; the snippets below assume that. To run
the same calls against the sandbox while exploring, prefix the session
with `dryad_use_sandbox()`.

```{r}
library(dryaddata)
```

## Query syntax

The `q` argument accepts the same shorthand the API does:

* Multi-term queries are AND-ed: `q = "soil carbon"` returns datasets
  matching *both* terms.
* Prefix a term with `-` to negate: `q = "carbon -ocean"`.
* Append `*` for prefix matching: `q = "soil*"`.

```{r}
dryad_search(q = "soil carbon", per_page = 3)
```

## Filtering

The named arguments map straight onto the underlying parameters:

```{r}
dryad_search(
  q                = "carbon",
  subject          = "soil science",
  orcid            = "0000-0001-7146-8135",
  funder           = "https://ror.org/021nxhr62",
  published_since  = "2024-01-01T00:00:00Z",
  published_before = "2026-12-31T23:59:59Z",
  per_page         = 5
)
```

`dryad_datasets()` exposes a similar set of filters but without free text
(`publication_issn`, `publication_name`, `manuscript_number`,
`curation_status`).

## Pagination

By default you get one page (`per_page` defaults to 20, max 100). For
larger result sets either step manually:

```{r}
page2 <- dryad_search(q = "soil", page = 2, per_page = 50)
```

or pull everything with `all_pages = TRUE`. The result combines the
underlying pages into a single `records` / `data` block; `metadata$pages`
tells you how many requests were made.

```{r}
all_soil <- dryad_search(q = "soil", per_page = 100, all_pages = TRUE)
nrow(all_soil$data)
all_soil$metadata$pages
```

Use `max_pages` as a circuit breaker for unfamiliar queries:

```{r}
dryad_search(q = "soil", per_page = 100, all_pages = TRUE, max_pages = 5)
```

## Working with nested fields

Because `data` keeps complex fields as list-columns, vectorized access is
a single `vapply()` or `lapply()` away. For example, pulling the primary
author from each hit:

```{r}
hits <- dryad_search(q = "soil", per_page = 5)
vapply(hits$data$authors, function(a) {
  if (length(a)) paste(a[[1]]$firstName, a[[1]]$lastName) else NA_character_
}, character(1))
```

When you need a fully-flattened representation, the raw `records` list is
trivial to hand to `tidyr::unnest_wider()` or `purrr::map_dfr()`.
